Spatiotemporal model of `r paste(params$species)` occupancy

### 2020-11-25 analysis/sand-lance/occupancy-model.Rmd
draft sand lance models

---
title: "Spatiotemporal model of `r paste(params$species)` biomass"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Sand Lance"
    region: "Non-HG surveys"
    knots: 400
    # covs: "-sand-depth-roms"
    # covs: "-depth-roms"
    # covs: "-sand-depth-temp"
    covs: "-mud-depth-temp"
    REML: FALSE
    AR1: TRUE
    fixed_spatial: TRUE
    update_model: TRUE
    # update_model: FALSE
    update_model_check: TRUE
    update_predictions: TRUE
---

@@

@@

```{r formula}
if(params$covs == "-sand-depth-roms"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(roms_scaled, k = 3)"
  ))
clim_cov <- "roms"
}

if(params$covs == "-depth-roms"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(depth_scaled, k = 3) + 
  s(roms_scaled, k = 3)"
  ))
clim_cov <- "roms"
}

if(params$covs == "-depth-roms"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(depth_scaled, k = 3) + 
  s(roms_scaled, k = 3)"
  ))
clim_cov <- "roms"
}

if(params$covs == "-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "density ~ 0 + as.factor(year_pair) + 
  s(muddy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

@@

paste("region =", region)
paste("climate covariate =", clim_cov)
paste("model label =", covs)
paste("knots =", knots)
paste("REML =", REML)
paste("AR1 =", AR1)
```

@@

# # rename raw_depth for plots 
# d <- d %>% mutate(depth = raw_depth) %>%
#   filter(ssid %in% model_ssid)

@@

    total_biomass <- sdmTMB::sdmTMB(
      data = data, mod.formula,
      # time_varying = ~ 0 + depth_scaled + depth_scaled2, 
      time = "year_pair", spde = spde,
      family = tweedie(link = "log"),
      ar1_fields = params$AR1,
      reml = params$REML,
      include_spatial = params$fixed_spatial
    )

    saveRDS(total_biomass, file = paste0(
      "models/", spp, "/mod-tot-biomass-", spp, 
      covs, "-", ssid_string, "-ar1-", params$AR1, "-reml-", params$REML, params$knots, ".rds"
    ))
}
```

@@

m <- readRDS(paste0(
  "models/", spp, "/mod-tot-biomass-", spp, 
  covs, "-", ssid_string, "-ar1-", params$AR1, 
  "-reml-", params$REML, params$knots, ".rds"
  ))

@@

saveRDS(point_predictions, file = paste0(
  "data/", spp, "/check-mod-predictions-", spp, 
  covs, "-", ssid_string, "-ar1-", params$AR1, "-reml-", 
  params$REML, params$knots, ".rds"
))
}
```

```{r}
pred <- readRDS(paste0(
  "data/", spp, "/check-mod-predictions-", spp, 
  covs, "-", ssid_string, "-ar1-", params$AR1, "-reml-", params$REML, params$knots, ".rds"
))

@@

```{r}
g <- ggplot(pred, 
    aes(density, residuals, colour = as.factor(ssid))) +
  geom_point() + scale_x_continuous(trans = "log10") +  
  facet_wrap(~year_pair)
g
```

```{r}
g1 <- ggplot(pred,
      aes(density, exp(est), colour = as.factor(ssid))) +
    geom_abline(slope=1, intercept = 0) +
    scale_x_continuous(trans = 'log10') +
    scale_y_continuous(trans = 'log10') +
    geom_point(alpha = 0.2) 
g1
```

```{r}
g2 <- ggplot(pred, 
      aes((est), residuals, colour = density)) +
    geom_point(alpha = 0.2) + # scale_x_continuous(trans = 'log10') +
    facet_wrap(~year)
g2
```

@@

### Model summary
```{r }
if(!exists("m")){
m <- readRDS(paste0(
    "models/", spp, "/mod-tot-biomass-", spp, 
      covs, "-", ssid_string, "-ar1-", params$AR1, "-reml-", params$REML, params$knots, ".rds"
  ))
}
m
max(m$gradients)
```

@@

(g_depth <- ggplot(p_depth, aes(depth, exp(est), 
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  coord_cartesian(ylim = c(0, max(exp(p_depth$est))*20), xlim = c(0, 150)) +
  geom_line() + geom_ribbon(alpha = 0.4))

ggsave(paste0("figs/depth-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_depth})
```

@@

(g_roms <- ggplot(p_roms, aes(roms, exp(est), 
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  coord_cartesian(ylim = c(0, max(exp(p_roms$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

ggsave(paste0("figs/roms-", spp, covs, ".png"), width = 5, height = 5)
}

try({g_roms})
```

```{r}
if(clim_cov == "temp"){
nd_temp <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = seq(min(data$temp_scaled), 
    max(data$temp_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_temp <- predict(m, newdata = nd_temp, se_fit = TRUE, re_form = NA)
p_temp$temp <- (p_temp$temp_scaled*data$temp_sd[1])+ data$temp_mean[1]

(g_temp <- ggplot(p_temp, aes(temp, exp(est), 
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0, max(exp(p_temp$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

ggsave(paste0("figs/temp-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_temp})
```

```{r}
if(covs %in% c("-sand-depth-roms", "-sand-depth-temp")){
nd_sand <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  sandy_scaled = seq(min(data$sandy_scaled), 
    max(data$sandy_scaled), length.out = 50),
  muddy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

(g_sand <- ggplot(p_sand, aes(sandy, exp(est),
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0,max(exp(p_sand$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

# ggplot(p_sand, aes(sandy, exp(est),
#   ymin = exp(est - 1.96 * est_se), 
#   ymax = exp(est + 1.96 * est_se))) +
#   coord_cartesian(ylim = c(0, max(exp(p_sand$est))*2), xlim = c(0,1)) +
#   geom_line() + geom_ribbon(alpha = 0.4)

ggsave(paste0("figs/sand-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_sand})
```

```{r}
if(covs %in% c("-mud-depth-roms", "-mud-depth-temp")){
nd_mud <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  muddy_scaled = seq(min(data$muddy_scaled), 
    max(data$muddy_scaled), length.out = 50),
  sandy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

(g_mud <- ggplot(p_mud, aes(muddy, exp(est),
  ymin = exp(est - 1.96 * est_se), 
  ymax = exp(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0,max(exp(p_mud$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

# ggplot(p_mud, aes(muddy, exp(est),
#   ymin = exp(est - 1.96 * est_se), 
#   ymax = exp(est + 1.96 * est_se))) +
#   coord_cartesian(ylim = c(0, max(exp(p_mud$est))*2), xlim = c(0,1)) +
#   geom_line() + geom_ribbon(alpha = 0.4)

ggsave(paste0("figs/mud-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_mud})
```

@@

  m <- readRDS(paste0("models/", spp, "/mod-tot-biomass-", spp, 
    covs, "-", ssid_string, "-ar1-", params$AR1, "-reml-", params$REML, params$knots, ".rds"
  ))

@@

  saveRDS(predicted, file = paste0(
    "data/", spp, "/predictions-", spp, 
    covs, "-", ssid_string, "-tot-biomass-ar1-", params$AR1, "-reml-", params$REML, params$knots, ".rds"
  ))
}
```

Transform estimates to biomass density in kg/ha
```{r}
rm(predictions)
rm(predicted)

  try({
    predicted <- readRDS(paste0("data/", spp, "/predictions-", spp, 
      covs, "-", ssid_string, "-tot-biomass-ar1-", 
      params$AR1, "-reml-", params$REML, params$knots, ".rds"
    ))
  })

  predictions <- predicted$data
  predictions$total_bio <- exp(predictions$est) 
  max_raster <- quantile(predictions$total_bio, 0.999)
  max_bio <- signif(quantile(predictions$total_bio, 1), digits = 2)

  predictions$prop_max <- predictions$total_bio/max_raster

  saveRDS(predictions, file = paste0(
    "data/", spp,
    "/just-predictions-", spp, covs, "-total-biomass-ar1-", 
    params$AR1, "-reml-", params$REML, params$knots, ".rds"
  ))
```

@@

p_mean <- predictions  %>% group_by(X, Y) %>%
  mutate(mean_est = mean(exp(est))) %>% 
  filter(year_pair == 2011) %>% 
  mutate(year = "all years") %>%
  plot_facet_map("mean_est",
    raster_limits = c(0, max_raster),
    transform_col = fourth_root_power,
    legend_position = legend_coords
  ) + labs(fill = "kg/ha") +
  ggtitle(paste0("Estimated biomass density of ", species, 
    " \n(max = ", max_bio, " kg/ha)"))
print(p_mean)
ggsave(paste0("figs/map-", spp, covs, "-mean-biomass.png"), width = 5, height = 5)
```

@@

ggsave(paste0("figs/map-", spp, covs, "-spatial-rf.png"), 
  width = 5, height = 5)
```

@@

print(p_st)
ggsave(paste0("figs/map-", spp, covs, "-spatiotemporal-rf.png"), 
  width = 5, height = 5)
```

>>>>

---
title: "Spatiotemporal model of `r paste(params$species)` occupancy"
author: "Philina English"
date: "`r Sys.Date()`"
output: html_document
params: 
    species: "Pacific Sand Lance"
    region: "Non-HG surveys"
    knots: 400
    # covs: "-sand-depth-roms"
    # covs: "-depth-roms"
    # covs: "-sand-depth-temp"
    # covs: "-mud-depth-temp"
    # covs: "-sand-mud-depth-temp"
    covs: "-sand-by-depth-mud-temp"
    REML: FALSE
    AR1: TRUE
    fixed_spatial: TRUE
    update_model: TRUE
    # update_model: FALSE
    update_model_check: TRUE
    update_predictions: TRUE
---

@@

logit2prob <- function(logit){
  odds <- exp(logit)
  prob <- odds / (1 + odds)
  return(prob)
}

@@

```{r formula}
if(params$covs == "-sand-depth-roms"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(roms_scaled, k = 3)"
  ))
clim_cov <- "roms"
}

if(params$covs == "-depth-roms"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(depth_scaled, k = 3) + 
  s(roms_scaled, k = 3)"
  ))
clim_cov <- "roms"
}

if(params$covs == "-depth-roms"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(depth_scaled, k = 3) + 
  s(roms_scaled, k = 3)"
  ))
clim_cov <- "roms"
}

if(params$covs == "-sand-depth-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(muddy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-sand-mud-depth-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  s(sandy_scaled, k = 3) + 
  s(muddy_scaled, k = 3) + 
  s(depth_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

if(params$covs == "-sand-by-depth-mud-temp"){
mod.formula <- as.formula(paste0(
  "present ~ 0 + as.factor(year_pair) + 
  te(sandy_scaled, depth_scaled, k = 5) + 
  s(muddy_scaled, k = 3) + 
  s(temp_scaled, k = 3)"
  ))
clim_cov <- "temp"
}

@@

paste("region =", region)
paste("climate covariate =", clim_cov)
paste("model label =", covs)
paste("knots =", knots)
paste("REML =", REML)
paste("AR1 =", AR1)
paste("formula =", mod.formula)
```

@@

@@

    presence_mod <- sdmTMB::sdmTMB(
      data = data, mod.formula,
      # time_varying = ~ 0 + depth_scaled + depth_scaled2, 
      time = "year_pair", spde = spde,
      family = binomial(link = "logit"),
      ar1_fields = params$AR1,
      reml = params$REML,
      include_spatial = params$fixed_spatial
    )

    saveRDS(presence_mod, file = paste0(
      "models/", spp, "/mod-tot-binomial-", spp, 
      covs, "-", ssid_string, "-ar1-", params$AR1, "-reml-", params$REML, params$knots, ".rds"
    ))
}
```

@@

m <- readRDS(paste0(
  "models/", spp, "/mod-tot-binomial-", spp, 
  covs, "-", ssid_string, "-ar1-", params$AR1, 
  "-reml-", params$REML, params$knots, ".rds"
  ))

@@

saveRDS(point_predictions, file = paste0(
  "data/", spp, "/check-binomial-mod-predictions-", spp, 
  covs, "-", ssid_string, "-ar1-", params$AR1, "-reml-", 
  params$REML, params$knots, ".rds"
))
}
```

```{r}
pred <- readRDS(paste0(
  "data/", spp, "/check-binomial-mod-predictions-", spp, 
  covs, "-", ssid_string, "-ar1-", params$AR1, "-reml-", params$REML, params$knots, ".rds"
))

@@

```{r}
g1 <- ggplot(pred,
      aes(logit2prob(est), present)) +
      # aes(as.factor(present), logit2prob(est))) +
    geom_smooth(method = "glm", 
    method.args = list(family = "binomial"), se=F)+
    geom_point(alpha = 0.2) 
g1
```

```{r}
g2 <- ggplot(pred, 
      aes(est, residuals, colour = as.factor(present))) +
    geom_point(alpha = 0.2) + # scale_x_continuous(trans = 'log10') +
    facet_wrap(~year)
g2
```

```{r}
g3 <- ggplot(pred, 
      aes(log_depth, residuals, colour = as.factor(present))) +
    geom_point(alpha = 0.2) + # scale_x_continuous(trans = 'log10') +
    facet_wrap(~year)
g3
```

@@

### Model summary
```{r }
if(!exists("m")){
m <- readRDS(paste0(
    "models/", spp, "/mod-tot-binomial-", spp, 
      covs, "-", ssid_string, "-ar1-", params$AR1, "-reml-", params$REML, params$knots, ".rds"
  ))
}
m
max(m$gradients)
```

@@

(g_depth <- ggplot(p_depth, aes(depth, logit2prob(est), 
  ymin = logit2prob(est - 1.96 * est_se), 
  ymax = logit2prob(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0, max(logit2prob(p_depth$est))*20), xlim = c(0, 150)) +
  geom_line() + geom_ribbon(alpha = 0.4))

ggsave(paste0("figs/depth-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_depth})
```

@@

(g_roms <- ggplot(p_roms, aes(roms, logit2prob(est), 
  ymin = logit2prob(est - 1.96 * est_se), 
  ymax = logit2prob(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0, max(logit2prob(p_roms$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

ggsave(paste0("figs/roms-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_roms})
```

```{r}
if(clim_cov == "temp"){
nd_temp <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, # use min depth for estimated error at relevant depths
  sandy_scaled = 0,
  muddy_scaled = 0,
  temp_scaled = seq(min(data$temp_scaled), 
    max(data$temp_scaled), length.out = 50),
  year_pair = 2011 # a chosen year
)
p_temp <- predict(m, newdata = nd_temp, se_fit = TRUE, re_form = NA)
p_temp$temp <- (p_temp$temp_scaled*data$temp_sd[1])+ data$temp_mean[1]

(g_temp <- ggplot(p_temp, aes(temp, logit2prob(est), 
  ymin = logit2prob(est - 1.96 * est_se), 
  ymax = logit2prob(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0, max(logit2prob(p_temp$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

ggsave(paste0("figs/temp-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_temp})
```

```{r}
if(covs %in% c("-sand-depth-roms", "-sand-depth-temp", "-sand-mud-depth-temp", "-sand-by-depth-mud-temp")){
nd_sand <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  sandy_scaled = seq(min(data$sandy_scaled), 
    max(data$sandy_scaled), length.out = 50),
  muddy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

(g_sand <- ggplot(p_sand, aes(sandy, logit2prob(est),
  ymin = logit2prob(est - 1.96 * est_se), 
  ymax = logit2prob(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0,max(logit2prob(p_sand$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

# ggplot(p_sand, aes(sandy, logit2prob(est),
#   ymin = logit2prob(est - 1.96 * est_se), 
#   ymax = logit2prob(est + 1.96 * est_se))) +
#   coord_cartesian(ylim = c(0, max(logit2prob(p_sand$est))*2), xlim = c(0,1)) +
#   geom_line() + geom_ribbon(alpha = 0.4)

ggsave(paste0("figs/sand-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_sand})
```

```{r}
if(covs %in% c("-mud-depth-roms", "-mud-depth-temp", "-sand-mud-depth-temp", "-sand-by-depth-mud-temp")){
nd_mud <- data.frame(
  depth_scaled = min(data$depth_scaled)-0.2, 
  muddy_scaled = seq(min(data$muddy_scaled), 
    max(data$muddy_scaled), length.out = 50),
  sandy_scaled = 0,
  roms_scaled = 0,
  temp_scaled = 0,
  year_pair = 2011 # a chosen year
)

@@

(g_mud <- ggplot(p_mud, aes(muddy, logit2prob(est),
  ymin = logit2prob(est - 1.96 * est_se), 
  ymax = logit2prob(est + 1.96 * est_se))) +
  # coord_cartesian(ylim = c(0,max(logit2prob(p_mud$est))*20)) +
  geom_line() + geom_ribbon(alpha = 0.4))

# ggplot(p_mud, aes(muddy, logit2prob(est),
#   ymin = logit2prob(est - 1.96 * est_se), 
#   ymax = logit2prob(est + 1.96 * est_se))) +
#   coord_cartesian(ylim = c(0, max(logit2prob(p_mud$est))*2), xlim = c(0,1)) +
#   geom_line() + geom_ribbon(alpha = 0.4)

ggsave(paste0("figs/mud-binomial-", spp, covs, ".png"), width = 5, height = 5)
}
try({g_mud})
```

@@

  m <- readRDS(paste0("models/", spp, "/mod-tot-binomial-", spp, 
    covs, "-", ssid_string, "-ar1-", params$AR1, "-reml-", params$REML, params$knots, ".rds"
  ))

@@

  saveRDS(predicted, file = paste0(
    "data/", spp, "/predictions-", spp, 
    covs, "-", ssid_string, "-tot-binomial-ar1-", params$AR1, "-reml-", params$REML, params$knots, ".rds"
  ))
}
```

Transform estimates to probability of occupancy
```{r}
rm(predictions)
rm(predicted)

  try({
    predicted <- readRDS(paste0("data/", spp, "/predictions-", spp, 
      covs, "-", ssid_string, "-tot-binomial-ar1-", 
      params$AR1, "-reml-", params$REML, params$knots, ".rds"
    ))
  })

  predictions <- predicted$data
  predictions$total_prob <- logit2prob(predictions$est) 
  max_raster <- quantile(predictions$total_prob, 0.999)
  max_prob <- signif(quantile(predictions$total_prob, 1), digits = 2)

  predictions$prop_max <- predictions$total_prob/max_raster

  saveRDS(predictions, file = paste0(
    "data/", spp,
    "/just-predictions-", spp, covs, "-total-binomial-ar1-", 
    params$AR1, "-reml-", params$REML, params$knots, ".rds"
  ))
```

@@

p_mean <- predictions  %>% group_by(X, Y) %>%
  mutate(mean_est = mean(logit2prob(est))) %>% 
  filter(year_pair == 2011) %>% 
  mutate(year = "all years") %>%
  plot_facet_map("mean_est",
    raster_limits = c(0, max_raster),
    transform_col = fourth_root_power,
    legend_position = legend_coords
  ) + labs(fill = "prob") +
  ggtitle(paste0("Estimated biomass density of ", species, 
    " \n(max = ", max_prob, ")"))
print(p_mean)
ggsave(paste0("figs/map-", spp, covs, "-mean-distribution.png"), width = 5, height = 5)
```

@@

ggsave(paste0("figs/map-", spp, covs, "-binomial-spatial-rf.png"), 
  width = 5, height = 5)
```

@@

print(p_st)
ggsave(paste0("figs/map-", spp, covs, "-binomial-spatiotemporal-rf.png"), 
  width = 5, height = 5)
```
